---
title: "R Notebook"
output: html_notebook
---

load libraries
```{r}
library(tidyverse)
library(skimr)
library(here)
```


Read in csv
```{r}
here::here()
candies_long <- read_csv(here("clean_data/candies_long.csv"))
```
 
1. What is the total number of candy ratings given across the three years. (Number of candy ratings, not the number of raters. Don’t count missing values)


ANSWER: 623053 have been given over 3 years.

```{r}
total_ratings <- candies_long %>% select(candies_rating) %>% 
  filter(!is.na(candies_rating)) %>% 
 summarise(count = n())
   
```


2.  What was the average age of people who are going out trick or treating?

```{r}

going_out <- candies_long %>% 
  select(age, going_out) %>% 
  filter(going_out == "Yes") %>% 
  summarise(going_out_avg = round(mean((age), na.rm = TRUE)))
  
```

ANSWER: Average age for those who were going out to trick or treat was around 35.


3. What was the average age of people who are not going trick or treating?

ANSWER: Average age for those who were not going out to trick or treat was around 39.

```{r}

not_going_out <- candies_long %>% 
  select(going_out, age) %>% 
  filter(going_out == "No") %>% 
  summarise(going_out_avg = round(mean((age), na.rm = TRUE)))


not_going_out  
```


4. For each of joy, despair and meh, which candy bar received the most of these ratings?

For the purpose of the analysis the following candies are considered candy bars:
x100_grand_bar, reggie_jackson_bar, hershey_s_milk_chocolate, mr_goodbar, caramellos, chick_o_sticks_we_don_t_k, coffee_crisp, three_musketeers, hersheys_dark_chocol, nestle_crunch, whatchamacallit_bars, milky_way, heath_bar, kinder_happy_hippo, butterfinger, kit_kat, snickers, twix, any_full_sized_candy_bar, tolberone_something_or_ot



```{r}
unique(candies_long$candie_name)
```


4.1.The most 'joy' ratings received 'Kit-Kat bar'.

```{r}

joy_rating <- candies_long %>% 
  select(candie_name, candies_rating) %>% 
  filter(candies_rating == "JOY") %>% 
  group_by(candie_name) %>% 
  summarise(count = n()) %>% 
  arrange(desc(count))

joy_rating

```


4.2 The most 'despair' ratings received 'Chick o stick' bar.

```{r}
despair_rating <- candies_long %>% 
  select(candie_name, candies_rating) %>% 
  filter(candies_rating == "DESPAIR") %>% 
  group_by(candie_name) %>% 
  summarise(count = n()) %>% 
  arrange(desc(count))

despair_rating

```

4.3 The most 'meh' ratings received '100_grand_bar' bar.

```{r}

meh_rating <- candies_long %>% 
  select(candie_name, candies_rating) %>% 
  filter(candies_rating == "MEH") %>% 
  group_by(candie_name) %>% 
  summarise(count = n()) %>% 
  arrange(desc(count))
  

meh_rating

```


5. How many people rated Starburst as despair?

ANSWER: 1990 people rated Starburst as "despair"

```{r}
starburst_despair <- candies_long %>% 
  select(candies_rating, candie_name) %>% 
  filter(candie_name == "starburst") %>% 
  filter(candies_rating == "DESPAIR") %>% 
  count()
 
  
starburst_despair
```


6. For the next three questions, count despair as -1, joy as +1, and meh as 0.


```{r}

candies_numeric <- candies_long %>% mutate(candies_rating_num = case_when(
  candies_rating == "MEH" ~ 0,
  candies_rating == "DESPAIR" ~ -1,
  candies_rating == "JOY" ~ 1))
```

7.What was the most popular candy bar by this rating system for each gender in the dataset ?

7.1 Females preferred twix.

```{r}
females_preferred <- candies_numeric %>% select(gender, candies_rating_num, candie_name) %>%
  filter(gender == "Female") %>% 
  filter(candies_rating_num == 1) %>%
  group_by(candie_name) %>% 
  summarise(n = sum(candies_rating_num)) %>% 
  arrange(desc(n)) 
  

females_preferred

```
7.2 Males preferred Kit Kats.

```{r}
males_preferred <- candies_numeric %>% select(gender, candies_rating_num, candie_name) %>%
  filter(gender == "Male") %>% 
  filter(candies_rating_num == 1) %>%
  group_by(candie_name) %>% 
  summarise(n = sum(candies_rating_num)) %>% 
  arrange(desc(n)) 


males_preferred
```


8. What was the most popular candy bar in each year?

8.1 The most popular candy bar in 2015 was Kit Kat.

```{r}
popular_2015 <- candies_numeric %>% select(year, candies_rating_num, candie_name) %>%
  filter(year == 2015) %>% 
  filter(candies_rating_num == 1) %>%
  group_by(candie_name) %>% 
  summarise(n = sum(candies_rating_num)) %>% 
  arrange(desc(n))

```
8.2 The most popular candy bar in 2016 was Kit Kat.

```{r}
popular_2016 <- candies_numeric %>% select(year, candies_rating_num, candie_name) %>%
  filter(year == 2016) %>% 
  filter(candies_rating_num == 1) %>%
  group_by(candie_name) %>% 
  summarise(n = sum(candies_rating_num)) %>% 
  arrange(desc(n))


```
8.3 The most popular candy bar in 2017 was Kit Kat.

```{r}
popular_2017 <- candies_numeric %>% select(year, candies_rating_num, candie_name) %>%
  filter(year == 2017) %>% 
  filter(candies_rating_num == 1) %>%
  group_by(candie_name) %>% 
  summarise(n = sum(candies_rating_num)) %>% 
  arrange(desc(n))

popular_2017
```

9. What was the most popular candy bar by this rating for people in US, Canada, UK, and all other countries?

9.1 The most popular candy bar in US was Kit Kat.
```{r}
US_popular <- candies_numeric %>% select(country, candies_rating_num, candie_name) %>%
  filter(country == "US") %>% 
  filter(candies_rating_num == 1) %>%
  group_by(candie_name) %>% 
  summarise(n = sum(candies_rating_num)) %>% 
  arrange(desc(n))


```
9.2 The most popular candy bar in the UK were Milky Way and Twix. 

```{r}
UK_popular <- candies_numeric %>% select(country, candies_rating_num, candie_name) %>%
  filter(country == "The UK") %>% 
  filter(candies_rating_num == 1) %>%
  group_by(candie_name) %>% 
  summarise(n = sum(candies_rating_num)) %>% 
  arrange(desc(n))


```
9.3 The most popular candy bar in Canada was Kit Kat.

```{r}
Canada_popular <- candies_numeric %>% select(country, candies_rating_num, candie_name) %>%
  filter(country == "Canada") %>% 
  filter(candies_rating_num == 1) %>%
  group_by(candie_name) %>% 
  summarise(n = sum(candies_rating_num)) %>% 
  arrange(desc(n))


```

9.4  Snickers was preferred in countries other than Us, Canada and UK.

```{r}


Other_popular <- candies_numeric %>% select(country, candies_rating_num, candie_name) %>%
  filter(country != "Canada") %>% 
  filter(country != "US") %>% 
  filter(country != "The UK") %>% 
  filter(!is.na(country)) %>% 
  filter(candies_rating_num == 1) %>%
  group_by(candie_name) %>% 
  summarise(n = sum(candies_rating_num)) %>% 
  arrange(desc(n))
```

