---
title: "R Notebook"
output: html_notebook
---

load libraries
```{r}
library(tidyverse)
library(skimr)
library(here)
```


Read in csv
```{r}
here::here()
candies <- read_csv(here("clean_data/candies.csv"))
```
 
1. What is the total number of candy ratings given across the three years. (Number of candy ratings, not the number of raters. Don’t count missing values)

```{r}
candies %>% 
  select(-c(year, age, going_out, country, gender)) %>% 
  summarise(across(.fns = ~sum(!is.na(.)))) %>% 
  summarise(rowSums(.))
  
```

ANSWER: 624769 have been given over 3 years.

2.  What was the average age of people who are going out trick or treating?

```{r}

going_out <- candies %>% 
  select(going_out, age) %>% 
  drop_na(age) %>% 
  filter(going_out == "Yes") %>% 
  summarise(going_out_avg = round(mean(age)))

going_out 

```
ANSWER: Average age for those who were going out to trick or treat was around 35.

3. What was the average age of people who are not going trick or treating?
```{r}

not_going_out <- candies %>% 
  select(going_out, age) %>% 
  drop_na(age) %>% 
  filter(going_out == "No") %>% 
  summarise(not_going_out_avg = round(mean(age)))

not_going_out  
```
ANSWER: Average age for those who were not going out to trick or treat was around 40.


4. For each of joy, despair and meh, which candy bar received the most of these ratings?
```{r}

candy_bars <- candies %>% 
  select(x100_grand_bar,reggie_jackson_bar,hershey_s_milk_choco,mr_goodbar,caramellos,chick_o_sticks_we_do,coffee_crisp,three_musketeers,hersheys_dark_chocol,nestle_crunch,whatchamacallit_bars,milky_way,heath_bar,kinder_happy_hippo,butterfinger,kit_kat,snickers,twix)
    
  

joy_rating <- colSums(candy_bars == "JOY", na.rm = TRUE) %>% 
  sort(decreasing = TRUE)

joy_rating


despair_rating <- colSums(candy_bars == "DESPAIR", na.rm = TRUE) %>% 
  sort(decreasing = TRUE)

despair_rating


meh_rating <- colSums(candy_bars == "MEH", na.rm = TRUE) %>% 
  sort(decreasing = TRUE)

meh_rating


```

ANSWER:
The most 'joy' ratings received 'Kit-Kat bar'.
The most 'despair' ratings received 'chick o stick' bar.
The most 'meh' ratings received '100_grand_bar' bar.

5. How many people rated Starburst as despair?
```{r}
starburst_despair <- candies %>% 
  select(starburst) %>%
  group_by(starburst) %>% 
  filter(starburst == "DESPAIR") %>% 
  count()

starburst_despair
```
ANSWER: 1990 people rated Starburst as "despair"

7. For the next three questions, count despair as -1, joy as +1, and meh as 0.

What was the most popular candy bar by this rating system for each gender in the dataset ?
What was the most popular candy bar in each year?
What was the most popular candy bar by this rating for people in US, Canada, UK, and all other countries?