---
title: "candies_project"
author: "J.Rosiak"
date: "2023-06-15"
output: html_document
---

## *TITLE: Candies*

The purpose of this project was to analyse some data on preferred and less preferred candies collected during Haloween trick or treating on three consecutive years: 2015 - 2017.

## *Project description*

There were three data sets with information from three years of Haloweeen Trick or Treating. The first part of the project involved preparing data for further analysis by transorming raw data into  clean file ready for further analysis.

The second part of the project involved using R to answer some questions related to candy ratings and people preferences about sweets collected depending on the year, country they live in and gender.

## *Software requirements*

Cleaning tasks and analysis were performed using RStudio. 

The following libraries were used:
```{r}
library(tidyverse)
library(skimr)
library(janitor)
library(here)
library(readxl)
library(stringr)
```


## *Assumptions*

For the purpose of the project I have made assumption on what candy is and used the definition offered by Collins Dictionary:
'candy is a sweet food, usually in small pieces or bars, made mainly from sugar or syrup, with flavoring, fruit, chocolate, nuts, etc. added.' I have excluded cakes from it.

I have also made an assumption that a candy bar 'is any form of confectionery in a solid usually rectangular block, especially one made with chocolate' - again following the definition offered by Collins Dictionary.

## *Data cleaning*
Data cleaning script is saved in the data_cleaning_scripts folder and is called cleaning_script_candies.R. Once the data has been read in I assigned each data set to a separate object.

**At first I worked with all three data sets simultaneously.**

Once the data was read in:

```{r}
here::here()
## 1.read in data
# 1.1
candy_2015 <- read_xlsx(here("..//task4/raw_data/boing-boing-candy-2015.xlsx"))

# 1.2
candy_2016 <- read_xlsx(here("..//task4/raw_data/boing-boing-candy-2016.xlsx"))

# 1.3
candy_2017 <- read_xlsx(here("../task4/raw_data/boing-boing-candy-2017.xlsx"))


## 2. Cleaned names

# 2.1
candy_2015 <- clean_names(candy_2015)

# 2.2
candy_2016 <- clean_names(candy_2016) 

# 2.3
candy_2017 <- clean_names(candy_2017) 



## 3.splitted time stamp column into 'year' and 'the rest'

# 3.1
candy_2015 <- candy_2015 %>% 
  separate(timestamp,c("year", "the_rest"), sep = "-" )

# 3.2
candy_2016 <- candy_2016 %>% 
  separate(timestamp,c("year", "the_rest"), sep = "-" )

# 3.3 adding a year column to a data set
candy_2017 <- candy_2017 %>% 
  mutate(year = "2017")

# 3.3.1 remove 'q1:_' from each column name
colnames(candy_2017) = str_remove(names(candy_2017), "^[a-z0-9]+[_]")

# 3.3.2 clean names again

candy_2017 <- clean_names(candy_2017) 


# 4. assigned all the columns, I wanted to drop to an object called 'drop_year'(first identify what columns are removed because they're not considered candy - 'not_candy_year', and the ones that are irrelevant to the analysis - 'irrelevant_col_year')

# 4.1
not_candy_2015 <- c("vials_of_pure_high_fructose_corn_syrup_for_main_lining_into_your_vein", "box_o_raisins","generic_brand_acetaminophen", "glow_sticks", "broken_glow_stick","healthy_fruit","hugs_actual_physical_hugs",  "creepy_religious_comics_chick_tracts","kale_smoothie","lapel_pins","minibags_of_chips", "spotted_dick", "peterson_brand_sidewalk_chalk", "peanut_butter_jars", "trail_mix","vicodin", "white_bread", "whole_wheat_anything","cash_or_other_forms_of_legal_tender")

irrelevant_col_2015 <- c("the_rest", "which_state_province_county_do_you_live_in",
               "please_leave_any_remarks_or_comments_regarding_your_choices","please_list_any_items_not_included_above_that_give_you_joy","please_list_any_items_not_included_above_that_give_you_despair","guess_the_number_of_mints_in_my_hand","betty_or_veronica","check_all_that_apply_i_cried_tears_of_sadness_at_the_end_of","that_dress_that_went_viral_early_this_year_when_i_first_saw_it_it_was","fill_in_the_blank_taylor_swift_is_a_force_for","what_is_your_favourite_font","if_you_squint_really_hard_the_words_intelligent_design_would_look_like","fill_in_the_blank_imitation_is_a_form_of","please_estimate_the_degree_s_of_separation_you_have_from_the_following_celebrities_jk_rowling","please_estimate_the_degree_s_of_separation_you_have_from_the_following_celebrities_jj_abrams","please_estimate_the_degree_s_of_separation_you_have_from_the_following_celebrities_beyonce","please_estimate_the_degree_s_of_separation_you_have_from_the_following_celebrities_bieber","please_estimate_the_degree_s_of_separation_you_have_from_the_following_celebrities_kevin_bacon","please_estimate_the_degree_s_of_separation_you_have_from_the_following_celebrities_francis_bacon_1561_1626","which_day_do_you_prefer_friday_or_sunday","please_estimate_the_degrees_of_separation_you_have_from_the_following_folks_bruce_lee","please_estimate_the_degrees_of_separation_you_have_from_the_following_folks_jk_rowling","please_estimate_the_degrees_of_separation_you_have_from_the_following_folks_malala_yousafzai","please_estimate_the_degrees_of_separation_you_have_from_the_following_folks_thom_yorke","please_estimate_the_degrees_of_separation_you_have_from_the_following_folks_jj_abrams","please_estimate_the_degrees_of_separation_you_have_from_the_following_folks_hillary_clinton","please_estimate_the_degrees_of_separation_you_have_from_the_following_folks_donald_trump","please_estimate_the_degrees_of_separation_you_have_from_the_following_folks_beyonce_knowles",
"please_leave_any_witty_snarky_or_thoughtful_remarks_or_comments_regarding_your_choices",
"that_dress_that_went_viral_a_few_years_back_when_i_first_saw_it_it_was", "do_you_eat_apples_the_correct_way_east_to_west_side_to_side_or_do_you_eat_them_like_a_freak_of_nature_south_to_north_bottom_to_top","when_you_see_the_above_image_of_the_4_different_websites_which_one_would_you_most_likely_check_out_please_be_honest", "york_peppermint_patties_ignore"
)

drop_2015 <- c(not_candy_2015, irrelevant_col_2015)

# 4.2

not_candy_2016 <- c("boxo_raisins","broken_glow_stick","cash_or_other_forms_of_legal_tender", "chardonnay","creepy_religious_comics_chick_tracts","generic_brand_acetaminophen","glow_sticks", "healthy_fruit","hugs_actual_physical_hugs", "person_of_interest_season_3_dvd_box_set_not_including_disc_4_with_hilarious_outtakes","spotted_dick", "trail_mix","vials_of_pure_high_fructose_corn_syrup_for_main_lining_into_your_vein","vicodin", "white_bread","whole_wheat_anything","sandwich_sized_bags_filled_with_boo_berry_crunch","sweet_tarts","real_housewives_of_orange_county_season_9_blue_ray", "minibags_of_chips", "kale_smoothie"
)

irrelevant_col_2016 <- c("the_rest", "which_state_province_county_do_you_live_in", "please_list_any_items_not_included_above_that_give_you_joy",                       "please_list_any_items_not_included_above_that_give_you_despair",   "please_leave_any_witty_snarky_or_thoughtful_remarks_or_comments_regarding_your_choices", "guess_the_number_of_mints_in_my_hand","betty_or_veronica", "that_dress_that_went_viral_a_few_years_back_when_i_first_saw_it_it_was", "what_is_your_favourite_font", "please_estimate_the_degree_s_of_separation_you_have_from_the_following_celebrities_jk_rowling",        "please_estimate_the_degree_s_of_separation_you_have_from_the_following_celebrities_jj_abrams",         "please_estimate_the_degree_s_of_separation_you_have_from_the_following_celebrities_beyonce",           "please_estimate_the_degree_s_of_separation_you_have_from_the_following_celebrities_bieber",            "please_estimate_the_degree_s_of_separation_you_have_from_the_following_celebrities_kevin_bacon", "please_estimate_the_degree_s_of_separation_you_have_from_the_following_celebrities_francis_bacon_1561_1626", "which_day_do_you_prefer_friday_or_sunday", "do_you_eat_apples_the_correct_way_east_to_west_side_to_side_or_do_you_eat_them_like_a_freak_of_nature_south_to_north_bottom_to_top", "when_you_see_the_above_image_of_the_4_different_websites_which_one_would_you_most_likely_check_out_please_be_honest","york_peppermint_patties_ignore")
                         
drop_2016 <- c(not_candy_2016, irrelevant_col_2016)

# 4.3
not_candy_2017 <- c("boxo_raisins","broken_glow_stick","cash_or_other_forms_of_legal_tender", "chardonnay","creepy_religious_comics_chick_tracts","generic_brand_acetaminophen","glow_sticks", "healthy_fruit","hugs_actual_physical_hugs", "kale_smoothie", "person_of_interest_season_3_dvd_box_set_not_including_disc_4_with_hilarious_outtakes","spotted_dick", "trail_mix","vials_of_pure_high_fructose_corn_syrup_for_main_lining_into_your_vein","vicodin", "white_bread","whole_wheat_anything", "bonkers_the_board_game")
  
irrelevant_col_2017 <- c("id","state_province_county_etc","joy_other","despair_other","other_comments","dress","x114","day","media_daily_dish","media_science","media_espn","media_yahoo","coordinates_x_y", "abstained_from_m_ming"
)

drop_2017 <- c(not_candy_2017, irrelevant_col_2017)


## 5. keep only columns that seem to include candies and are relevant to the analysis

# 5.1 
candy_2015 <- candy_2015 %>% 
  select(!any_of(drop_2015))

# 5.2
candy_2016 <- candy_2016 %>% 
  select(!any_of(candy_2016))

# 5.3
candy_2017 <- candy_2017 %>% 
  select(!any_of(candy_2017))



## 6. ensure each data set contains year, age, going_out, country, gender and other column names are likely to match

# 6.1

candy_2015 <- candy_2015 %>% 
  rename("age" = "how_old_are_you",
         "going_out" = "are_you_going_actually_going_trick_or_treating_yourself",
         "anonymous_brown_globs" =
           "anonymous_brown_globs_that_come_in_black_and_orange_wrappers")
         

# 6.2

candy_2016 <- candy_2016 %>% 
  rename("age" = "how_old_are_you",
         "gender" = "your_gender",
         "going_out" = "are_you_going_actually_going_trick_or_treating_yourself","country" = "which_country_do_you_live_in",
         "anonymous_brown_globs" =
           "anonymous_brown_globs_that_come_in_black_and_orange_wrappers")

# 6.3 all required columns already there

candy_2017 <- candy_2017 %>% 
  rename("anonymous_brown_globs" =
  "anonymous_brown_globs_that_come_in_black_and_orange_wrappers_a_k_a_mary_janes")


## 7. bind all three data sets into one called 'candies' and explore further.

# 7.1.

candies <- bind_rows(candy_2015, candy_2016, candy_2017)

# 7.2 shorten column names to 25 characters for ease of analysis ( 25 characters chosen as it guaranteed inclusion of all candy bars and 'any_full_sized_candy_bar' column)

short_col_names <- str_sub(colnames(candies),start = 1, end = 25)

colnames(candies) <- short_col_names

# 8. remove values no longer needed in the script

rm(drop_2015, drop_2016, drop_2017)
rm(irrelevant_col_2015, irrelevant_col_2016,irrelevant_col_2017)
rm(not_candy_2015,not_candy_2016, not_candy_2017)

## 9. change age column to numeric and ensure that data provided falls between certain values

candies <- candies %>% 
  mutate(age = as.integer(age),
         age = if_else(age > 0 & age <= 123, age , NA))



## 10. tidy up a 'country' column

# 10.1 check what distinctive country names there are

countries_distinct <- candies %>% 
  select(country) %>% 
  distinct(country) %>% 
  arrange(country)


# 10.2 change everything to sentence case, remove any punctuation, replace countries that have a number as a value to 'US' and then replacing all potential entries that suggest UK as a country with 'UK' and 'Canada'.

candies <- candies %>% 
  mutate(country = gsub("[[:punct:]]", "", country),   # remove any punctuation
       country = str_to_sentence(country),
       country = str_replace_all(country,"[0-9]+", "US"), # change numbers to US based on info from a 'state_province' column. 
       country = case_when(country == "United kingdom" ~ "The UK",
                           country == "United kindom"  ~ "The UK",
                           country == "Uk"             ~ "The UK",
                           country == "England"        ~ "The UK",
                           country == "Endland"        ~ "The UK",
                           country == "Scotland"       ~ "The UK",
                           country == "Ud"             ~ NA,
                           country == "Unhinged states"~ NA,
                           country == "Can"            ~ "Canada",
                           country == "Canae"          ~ "Canada",
                           country == "Soviet canuckistan" ~ "Canada",
                           TRUE                        ~ country))
               
# 10.3 tidy up countries that seem to be 'USA'

candies <- candies %>% 
  mutate(country = str_replace(country, "^U[a-z ]+[a-z]+", "US"),
         country = case_when(
            country == "Us" ~ "US",
            country == "The united states" ~ "US",
            country == "The united states of america" ~ "US",
            country == "The yoo ess of aaayyyyyy" ~ "US",
            country == "Pittsburgh" ~ "US",
            country == "North carolina" ~ "US",
            country == "New jersey" ~ "US",
            country == "New york" ~ "US",
            country == "I pretend to be from canada but i am really from the united states"   ~ "US",
            country == "California" ~ "US",
            country == "Alaska"  ~ "US",
            country == "America" ~ "US",
            country == "I don't know anymore" ~ "US",
            country == "The best one  usa" ~ "US",
            country == "Trumpistan" ~ "US",
            country == "Narnia" ~ "US",
            country == "Ahemamerca" ~ "US",
            country == "A" ~ "US",
            country == "The best one  usa" ~ "US",
            country == "Murica" ~ "US",  
            country == "Murrika" ~ "US",
            country == "Merica" ~ "US",
            country == "Gods country" ~ "US",
            country == "N america" ~ "US",
            country == "Subcanadian north america merica" ~ "US",
            TRUE     ~ country
))

# 10.4 observations that do not allow to identify a specific country are being changed to 'NA'.


candies <- candies %>% 
  mutate(country = case_when(country == "Atlantis" ~ NA,
       country == "Denial" ~ NA,
       country == "Fear and loathing" ~ NA,
       country == "Insanity lately" ~ NA,
       country == "One of the best ones" ~ NA,
       country == "See above" ~ NA,
       country == "Somewhere" ~ NA,
       country == "Subscribe to dmUSuzUS on youtube" ~ NA,
       country == "The republic of cascadia" ~ NA, # could be either Canada or US
       country == "Cascadia" ~ NA,
       country == "There isnt one for old men" ~ NA,
       country == "This one" ~ NA,
       country == "Earth" ~ NA,
       country == "A tropical island south of the equator" ~ NA,
       country == "I dont know anymore" ~ NA,
       country == "Not the usa or canada" ~ NA,
       country == "Eua" ~ NA,
       country == "Europe" ~ NA,
       country == "Neverland" ~ NA,
       TRUE     ~ country
   ))	

# 10.5 change all character 'NA's to logical

candies <- candies %>% 
  mutate(country = case_when(country == "NA" ~ NA,
                             TRUE     ~ country))
                         
 
# 11. change the data frame format from wide to long to help with answering further analysis questions

# 11.1 assign all candy related part of df to columns_candy

columns_candy <- candies %>% select(-c(year, age, gender, country, going_out))

# 11.2 change the format to long and assign it to an object called "candies_long." 

candies_long <- candies %>% 
  pivot_longer(cols = names(columns_candy), 
               names_to = "candie_name", 
               values_to = "candies_rating")

# 11.3 write a new csv into a clean data folder called 'candies_long.'
here::here()

write_csv(candies_long, file = (here("clean_data/candies_long.csv")))



```

