---
title: "Decathlon"
author: "J.Rosiak"
date: "2023-06-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## *DECATHLON Project*

This project involved preparing decathlon data for further analysis and the analysis themselves using RStudio.

## *Overview*

Decathlon data focused on two main competitions - Olympic and Decastar and included the names of different sportsmen and how they performed in various competitions.

## *Tidying Up*

The following libraries have been used:

* library(tidyverse)

* library(janitor)

* library(here)


I have noticed that there were row names, so converted them into names column and reassigned to decathlon
```{r}

rownames_to_column(decathlon, var = " ")

has_rownames(decathlon)

decathlon <- rownames_to_column(decathlon, var = "names") %>% as_tibble()
```

Then cleaned the names using a janitor package and renamed columns

```{r}
decathlon <- rename(decathlon, "names" = "names", "short_run" = "x100m", "long_jump" = "long_jump", 
"shot_put" ="shot_put", "high_jump" = "high_jump", "long_run" = "x400m",
"hurdle_run" = "x110m_hurdle","dicus" = "discus", "pole_vault" = "pole_vault","javeline" = "javeline", "1500_run" = "x1500m", "rank" = "rank", "points" = "points","competition_name" = "competition")
```

Afterwards I have checked how many missing values were in a data set

```{r}
missing_values <- sum(is.na(decathlon))

```

Once I was happy with the data I saved it into a csv data file 'decathlon_clean.csv'.

As the data set looked small I have decided to keep it in the wide format as it was more pleasing to an eye.


## *Analysis*

Using RStudio and the following libraries:

* library(tidyverse)

* library(here)

### I have focused on answering the following questions:

**1. Who had the longest long jump seen in the data?**

   Clay had the longest jump (7.96).
   
```{r}
longest_jump <- decathlon %>% 
  select(names, long_jump) %>% 
  arrange(desc(long_jump)) %>% 
  head()

```


**2. What was the average 100m time in each competition?**

Decastar	11.17538

OlympicG	10.91571

```{r}
avg_100m <- decathlon %>% 
  select(short_run, competition_name) %>% 
  group_by(competition_name) %>% 
  summarise(avg_short_run = mean(short_run))
```

**3. Who had the highest total points across both competitions?**

Sebrle had the highest total score of	17110.

```{r}
highest_total_score <- decathlon %>%
  select(names, points, competition_name) %>% 
  group_by(names) %>% 
  summarise(final_score = sum(points)) %>% 
  arrange(desc(final_score)) %>% 
  head(n = 1)
```


**4. What was the shot-put scores for the top three competitors in each competition?**

Yurkov	15.19	Decastar		
Sebrle	14.83	Decastar		
Karpov	14.77	Decastar		
Sebrle	16.36	OlympicG		
Karpov	15.93	OlympicG		
Macey	  15.73	OlympicG

```{r}
shot_put_best <- decathlon %>% 
  select(names, shot_put, competition_name) %>% 
  group_by(competition_name)%>% 
  arrange(desc(shot_put)) %>%
  slice_max(shot_put, n = 3)
 
```

**5. What was the average points for competitors who ran the 400m in less than 50 seconds vs. those than ran 400m in more than 50 seconds?**

The average points for those who run 400m in less than 50 seconds was 7727, and the average points for those who run in more than 50s was 8120.

```{r}
avg_points_longrun <- decathlon %>% 
  select(long_run, points) %>% 
  mutate(above_50 = if_else(long_run > 50, TRUE, FALSE)) %>% 
  group_by(above_50) %>% 
  summarise(average_points = mean(points))
```

